String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-05-09 2023-1-19/sections/overview.tex
overview
some syntax error in overview

We first construct an automaton for the regular expression $(\Sigma \setminus a)^{\{1, 60\}} (\Sigma \setminus b)^{\{1, 60\}} (\Sigma \setminus c)^{\{0, 60\}}$. We introduce three registers, say $r_1, r_2, r_3$, to represent the three counting operators.  
%Moreover, we introduce another register, say $r_0$, to denote the length of strings. 
Then we first construct a nondeterministic finite automaton (NFA) for $(\Sigma \setminus a)^* (\Sigma \setminus b)^* (\Sigma \setminus c)^*$, then add the updates of registers to transitions, resulting into a so-called cost-enriched finite automaton (CEFA, see Section~\ref{sec:automaton} for the definition) $\aut_1$,  as illustrated in Figure~\ref{fig:overview}(a). 
%Then we obtain a cost-enriched finite automaton (CEFA) $\aut_1$ that is illustrated in Figure~\ref{fig:overview}(a), where the self-loops around $p_0, p_1, p_2$ represent the first, second, and third counting operator respectively, in which $r_1, r_2, r_3$ are incremented. 
%Moreover, $r_0$ is incremented in all the three self-loops. 
Furthermore, the counting bounds are specified by the accepting condition $1 \le r_1 \le 60 \wedge 1 \le r_2 \le 60 \wedge 0 \le r_3 \le 60$. 
A string $w$ is accepted by $\aut_1$ if when reading the characters in $w$, $\aut_1$ applies the transitions to update the state and the values of registers, and reaches a final state $q$ in the end, moreover, the resulting values of the three registers, say $v_1, v_2, v_3$, satisfy the accepting condition. 
Note here we represent the counting operators symbolically by registers, instead of unfolding them explicitly. 
Moreover, we construct another CEFA $\aut_2$ for $\Sigma^* c^+$ (see Figure~\ref{fig:overview}(b)), which is actually an NFA, since $\Sigma^* c^+$ contains no (explicit) counting operators. Finally, we construct the third CEFA $\aut_3$ (see Figure~\ref{fig:overview}(c)), where a register $r_4$ is used to denote the length of strings and the accepting condition is $\ltrue$. (See Section~\ref{subsec:regex2cefa} for more details about the construction of CEFA.)
%where the register $r_0$ is used to record the length of strings.

@@

Next, 
%we remove the $\varepsilon$-transitions of $\aut_1$, resulting into $\aut'_1$, 
we construct $\aut_1 \cap \aut_2 \cap \aut_3$, that is, the intersection (aka product) of $\aut_1$, $\aut_2$, and $\aut_3$, as illustrated in Figure~\ref{fig:overview:product}(a), where the states from which no final states are reachable are removed. 
%Furthermore, we add one special register, say $r_0$, to record the length of strings. Note that $r_0$ is incremented in all transitions. 
%In $\aut_1 \cap \aut_2 \cap \aut_3$, the accepting condition $1 \le r_1 \le 60 \wedge 1 \le r_2 \le 60 \wedge 0 \le r_3 \le 60 \wedge r_0 > 120$ is attached to both $(p_0,q_1)$ and $(p_1,q_1)$, where $r_0 > 120$ is added as a conjunct to express $|x| > 120$.  
For technical convenience, we also think the updates on the values of registers in transitions as vectors $(v_1, v_2, v_3, v_4)$, where $v_i \in \Int$ is the update on the register $r_{i}$ for each $i \in [4]$. For instance, the transitions corresponding to the self-loop around $(p_0, q_0, q'_0)$ are thought as $((p_0, q_0, q'_0), a', (p_0, q_0, q'_0), (1,0,0,1))$ with $a' \in \Sigma \setminus \{a\}$, since $r_1$ and $r_4$ are incremented in these transitions. With the updates thought as vectors, the resulting CEFA is illustrated in Figure~\ref{fig:overview:product}(b).
%the transitions corresponding to the self-loop is $(q, a', q', (1,1,0,0))$ with $a' \in \Sigma \setminus \{a\}$.

@@

%For instance, in the transition from $(p_0, q_1)$ to $(p_1, q_1)$, $r_0$ is incremented, and the values of the other registers is unchanged. Therefore, the vector for this transition is $(1, 0, 0, 0)$. 
The satisfiability of the original string constraint is reduced to the nonemptiness of the CEFA $\aut_1 \cap \aut_2 \cap \aut_3$ with respect to the LIA formula $\varphi \equiv r_4 > 120$, that is, whether there exist $w \in \Sigma^*$ and $(v_1, v_2, v_3, v_4) \in \Int^4$ such that $w$ is accepted by $\aut_1 \cap \aut_2 \cap \aut_3$, producing a cost $(v_1, v_2, v_3, v_4)$ that satisfies $1 \le v_1 \le 60\wedge 1 \le v_2 \le 60 \wedge 0 \le v_3 \le 60$ and $v_4 > 120$. 
It is not hard to observe that the nonemptiness of $\aut_1 \cap \aut_2 \cap \aut_3$ w.r.t. $\varphi \equiv r_4 > 120$ is independent of the characters of $\aut_1 \cap \aut_2 \cap \aut_3$.  Therefore, the characters in $\aut_1 \cap \aut_2 \cap \aut_3$ can be ignored, resulting into an NFA $\cB$ over the alphabet $\costset$ (see Figure~\ref{fig:overview:product:reduced}(a)), where $\costset$ is the set of vectors from $\Int^4$ occurring in the transitions of $\aut_1 \cap \aut_2 \cap \aut_3$. Then the nonemptiness of $\aut_1 \cap \aut_2 \cap \aut_3$ w.r.t. $\varphi \equiv r_4 > 120$ is reduced to the problem of deciding whether there exists $w' \in \costset^*$ such that it is accepted by $\cB$ and its Parikh image (i.e. numbers of occurrences of characters), say $\eta_{w'}: \costset \rightarrow \Nat$, satisfies that $1 \le v'_1 \le 60\wedge 1 \le v'_2 \le 60 \wedge 0 \le v'_3 \le 60$ and $v'_4 > 120$, where $\myvec{v'} = \sum \limits_{\myvec{v} \in \costset} \eta_{w'}(\myvec{v}) \myvec{v}$. Intuitively, $\myvec{v'}$ here is a weighted sum of vectors $\myvec{v} \in \costset$, where the weight for $\myvec{v}$ is the number of occurrences of $\myvec{v}$ in $w'$. (See Section~\ref{subsec:cefadec} for more detailed arguments.)

@@

Nevertheless, it turns out that the algorithm illustrated above is insufficient to solve the string constraints with regex-counting and string-length efficiently, since the sizes of NFA obtained from CEFA can still be big, resulting into the LIA formulas of big sizes, while the satisfiability of existential LIA formulas is NP-complete (see e.g. \cite{Haase18}). 
To tackle this issue, we proposes techniques to reduce the sizes of NFA over the alphabet $\costset$.

@@

%Our main idea to reduce the size of the CEFA $\aut_1 \times \aut_2$ is to \emph{ignore the characters in transitions}, since the updates of the values of registers in transitions are independent of the characters therein and the satisfiability of the formula in~(\ref{eqn-LIA}) only concerns about the values of registers. After the characters in transitions are ignored, each transition in $\aut_1 \times \aut_2$ is of the form $(q, \myvec{v}, q')$, where $\myvec{v}$ is the vector denoting the updates on the values of registers. Let $\cB$ denote the resulting automaton. We then think $\cB$ as an NFA, where the vectors $\myvec{v}$ are taken as the characters.

We determinize $\cB$, apply the minimization algorithm to the resulting deterministic finite automaton (DFA), and obtain a DFA $\cC$, as illustrated in Figure~\ref{fig:overview:product:reduced}(b). 
Note that $\cC$ contains only three states $q''_0, q''_1, q''_2$ and six transitions, while $\cB$ contains four states and eight transitions. 
Although the determinization and minimization may not reduce the sizes of NFA in the worst case, the experimental results in Section~\ref{sec:implementation} show us that they do reduce the sizes of NFA that are constructed from the regular expressions in practice. 
%thus containing much less number of transitions than $\aut_1 \times \aut_2$\footnote{This is the case even if we consider subsets of characters, e.g. $\{a\}, \{b\}, \{c\}, \Sigma \setminus \{a, b, c\}$, in the transitions of $\aut_1 \times \aut_2$.}.
%Then we reinterpret $\cC$ as a CEFA with the acceptance condition $\alpha(r_0, r_1, r_2, r_3)$. 
%Finally, we compute an LIA formula $\psi'(i_1, \cdots, i_6)$ from $\cC$ and reduce the satisfiability of the string constraint in~(\ref{eqn-running}) to the satisfiability of the following existential LIA formula
%\begin{equation}\label{eqn-LIA-reduced}
%\psi'(i_1, \cdots, i_6) \wedge \bigwedge \limits_{0 \le j \le 3} r_j = \sum \limits_{k \in [6]}  i_k v'_{k, j+1} \wedge \alpha(r_0, r_1, r_2, r_3),
%\end{equation}
%where $\myvec{v'_{k}}=(v'_{k,1}, v'_{k,2}, v'_{k,3}, v'_{k,4})$ is the vector of updates of registers of the transition $\tau'_k$ for $k \in [6]$.

@@

At last, for the string constraints that are satisfied by a collection of strings whose lengths are small, computing LIA formulas from NFA and solving them by SMT solvers are still expensive.  
Therefore, we introduce a heuristic of searching for solutions based on \emph{under approximations}, before computing LIA formulas and solving them by SMT solvers. More precisely, we enforce a constant upper bound $B$ on the lengths of strings and search for a $B$-bounded satisfiable assignment, that is, a satisfiable assignment where the lengths of strings are bound by $B$. Initially we set $\ell$, an integer variable to denoting the lengths of strings, to $1$. If there does not exist a $\ell$-bounded satisfiable assignment, then we let $\ell:=\ell+1$ and continue the search, until finding a satisfiable assignment or reaching the bound $B$. For instance, if we remove $|x| > 120$ from the string constraint~(\ref{eqn-running}), then the resulting constraint is satisfiable. Setting the bound $B$ to $3$ is sufficient to find a solution, which is witnessed by the following accepting run of $\cC$: $q^{\prime\prime}_0 \xrightarrow{(1,0,0,1)} q^{\prime\prime}_0 \xrightarrow{(0,1,0,1)} q^{\prime\prime}_1 \xrightarrow{(0,0,0,1)} q^{\prime\prime}_2$. Note that the accepting run produces $(1,1,0,3)$, a vector of register values,  which satisfies $1 \le r_1 \le 60 \wedge 1 \le r_2 \le 60 \wedge 0 \le r_3 \le 60$.

>>>>

We first construct an automaton for the regular expression $(\Sigma \setminus a)^{\{1, 60\}} (\Sigma \setminus b)^{\{1, 60\}} (\Sigma \setminus c)^{\{0, 60\}}$. We introduce three registers, say $r_1, r_2, r_3$, to represent the three counting operators.  
%Moreover, we introduce another register, say $r_0$, to denote the length of strings. 
Then we first construct a nondeterministic finite automaton (NFA) for $(\Sigma \setminus a)^* (\Sigma \setminus b)^* (\Sigma \setminus c)^*$, then add the updates of registers to transitions, resulting into a so-called cost-enriched finite automaton (CEFA, see Section~\ref{sec:automaton} for the definition) $\aut_1$,  as illustrated in Figure~\ref{fig:overview}(a). 
%Then we obtain a cost-enriched finite automaton (CEFA) $\aut_1$ that is illustrated in Figure~\ref{fig:overview}(a), where the self-loops around $p_0, p_1, p_2$ represent the first, second, and third counting operator respectively, in which $r_1, r_2, r_3$ are incremented. 
%Moreover, $r_0$ is incremented in all the three self-loops. 
Furthermore, the counting bounds are specified by the accepting condition $1 \le r_1 \le 60 \wedge 1 \le r_2 \le 60 \wedge 0 \le r_3 \le 60$. 
A string $w$ is accepted by $\aut_1$ if\denghang{,} when reading the characters in $w$, $\aut_1$ applies the transitions to update the state and the values of registers, and reaches a final state $q$ in the end, moreover, the resulting values of the three registers, say $v_1, v_2, v_3$, satisfy the accepting condition. 
Note here\denghang{Note that} we represent the counting operators symbolically by registers, instead of unfolding them explicitly. 
Moreover, we construct another CEFA $\aut_2$ for $\Sigma^* c^+$ (see Figure~\ref{fig:overview}(b)), which is actually an NFA, since $\Sigma^* c^+$ contains no (explicit) counting operators. Finally, we construct the third CEFA $\aut_3$ (see Figure~\ref{fig:overview}(c)), where a register $r_4$ is used to denote the length of strings and the accepting condition is $\ltrue$. (See Section~\ref{subsec:regex2cefa} for more details about the construction of CEFA.)
%where the register $r_0$ is used to record the length of strings.

@@

Next, 
%we remove the $\varepsilon$-transitions of $\aut_1$, resulting into $\aut'_1$, 
we construct $\aut_1 \cap \aut_2 \cap \aut_3$, that is, the intersection (aka product) of $\aut_1$, $\aut_2$, and $\aut_3$, as illustrated in Figure~\ref{fig:overview:product}(a), where the states from which no final states are reachable are removed. 
%Furthermore, we add one special register, say $r_0$, to record the length of strings. Note that $r_0$ is incremented in all transitions. 
%In $\aut_1 \cap \aut_2 \cap \aut_3$, the accepting condition $1 \le r_1 \le 60 \wedge 1 \le r_2 \le 60 \wedge 0 \le r_3 \le 60 \wedge r_0 > 120$ is attached to both $(p_0,q_1)$ and $(p_1,q_1)$, where $r_0 > 120$ is added as a conjunct to express $|x| > 120$.  
For technical convenience, we also think \denghang{of} the updates on the values of registers in transitions as vectors $(v_1, v_2, v_3, v_4)$, where $v_i \in \Int$ is the update on the register $r_{i}$ for each $i \in [4]$. For instance, the transitions corresponding to the self-loop around $(p_0, q_0, q'_0)$ are thought as $((p_0, q_0, q'_0), a', (p_0, q_0, q'_0), (1,0,0,1))$ with $a' \in \Sigma \setminus \{a\}$, since $r_1$ and $r_4$ are incremented in these transitions. With the updates thought \denghang{of} as vectors, the resulting CEFA is illustrated in Figure~\ref{fig:overview:product}(b).
%the transitions corresponding to the self-loop is $(q, a', q', (1,1,0,0))$ with $a' \in \Sigma \setminus \{a\}$.

@@

%For instance, in the transition from $(p_0, q_1)$ to $(p_1, q_1)$, $r_0$ is incremented, and the values of the other registers is unchanged. Therefore, the vector for this transition is $(1, 0, 0, 0)$. 
The satisfiability of the original string constraint is reduced to the nonemptiness of the CEFA $\aut_1 \cap \aut_2 \cap \aut_3$ with respect to the LIA formula $\varphi \equiv r_4 > 120$, that is, whether there exist\denghang{exist to exists} $w \in \Sigma^*$ and $(v_1, v_2, v_3, v_4) \in \Int^4$ such that $w$ is accepted by $\aut_1 \cap \aut_2 \cap \aut_3$, producing a cost $(v_1, v_2, v_3, v_4)$ that satisfies $1 \le v_1 \le 60\wedge 1 \le v_2 \le 60 \wedge 0 \le v_3 \le 60$ and $v_4 > 120$. 
It is not hard to observe that the nonemptiness of $\aut_1 \cap \aut_2 \cap \aut_3$ w.r.t. $\varphi \equiv r_4 > 120$ is independent of the characters of $\aut_1 \cap \aut_2 \cap \aut_3$.  Therefore, the characters in $\aut_1 \cap \aut_2 \cap \aut_3$ can be ignored, resulting into an NFA $\cB$ over the alphabet $\costset$ (see Figure~\ref{fig:overview:product:reduced}(a)), where $\costset$ is the set of vectors from $\Int^4$ occurring in the transitions of $\aut_1 \cap \aut_2 \cap \aut_3$. Then the nonemptiness of $\aut_1 \cap \aut_2 \cap \aut_3$ w.r.t. $\varphi \equiv r_4 > 120$ is reduced to the problem of deciding whether there exists $w' \in \costset^*$ such that it is accepted by $\cB$ and its Parikh image (i.e. numbers of occurrences of characters), say $\eta_{w'}: \costset \rightarrow \Nat$, satisfies that $1 \le v'_1 \le 60\wedge 1 \le v'_2 \le 60 \wedge 0 \le v'_3 \le 60$ and $v'_4 > 120$, where $\myvec{v'} = \sum \limits_{\myvec{v} \in \costset} \eta_{w'}(\myvec{v}) \myvec{v}$. Intuitively, $\myvec{v'}$ here is a weighted sum of vectors $\myvec{v} \in \costset$, where the weight for $\myvec{v}$ is the number of occurrences of $\myvec{v}$ in $w'$. (See Section~\ref{subsec:cefadec} for more detailed arguments.)

@@

Nevertheless, it turns out that the algorithm illustrated above is insufficient to solve the string constraints with regex-counting and string-length efficiently, since the sizes of NFA obtained from CEFA can still be big, resulting into\denghang{result in} the LIA formulas of big sizes, while the satisfiability of existential LIA formulas is NP-complete (see e.g. \cite{Haase18}). 
To tackle this issue, we proposes techniques to reduce the sizes of NFA over the alphabet $\costset$.

@@

%Our main idea to reduce the size of the CEFA $\aut_1 \times \aut_2$ is to \emph{ignore the characters in transitions}, since the updates of the values of registers in transitions are independent of the characters therein and the satisfiability of the formula in~(\ref{eqn-LIA}) only concerns about the values of registers. After the characters in transitions are ignored, each transition in $\aut_1 \times \aut_2$ is of the form $(q, \myvec{v}, q')$, where $\myvec{v}$ is the vector denoting the updates on the values of registers. Let $\cB$ denote the resulting automaton. We then think \denghang{of} $\cB$ as an NFA, where the vectors $\myvec{v}$ are taken as the characters.

We determinize $\cB$, apply the minimization algorithm to the resulting deterministic finite automaton (DFA), and obtain a DFA $\cC$, as illustrated in Figure~\ref{fig:overview:product:reduced}(b). 
Note that $\cC$ contains only three states $q''_0, q''_1, q''_2$ and six transitions, while $\cB$ contains four states and eight transitions. 
Although the determinization and minimization may not reduce the sizes of NFA in the worst case, the experimental results in Section~\ref{sec:implementation} show us that they do reduce the sizes of NFA that are \denghang{too wordy, remove that are} constructed from the regular expressions in practice. 
%thus containing much less number of transitions than $\aut_1 \times \aut_2$\footnote{This is the case even if we consider subsets of characters, e.g. $\{a\}, \{b\}, \{c\}, \Sigma \setminus \{a, b, c\}$, in the transitions of $\aut_1 \times \aut_2$.}.
%Then we reinterpret $\cC$ as a CEFA with the acceptance condition $\alpha(r_0, r_1, r_2, r_3)$. 
%Finally, we compute an LIA formula $\psi'(i_1, \cdots, i_6)$ from $\cC$ and reduce the satisfiability of the string constraint in~(\ref{eqn-running}) to the satisfiability of the following existential LIA formula
%\begin{equation}\label{eqn-LIA-reduced}
%\psi'(i_1, \cdots, i_6) \wedge \bigwedge \limits_{0 \le j \le 3} r_j = \sum \limits_{k \in [6]}  i_k v'_{k, j+1} \wedge \alpha(r_0, r_1, r_2, r_3),
%\end{equation}
%where $\myvec{v'_{k}}=(v'_{k,1}, v'_{k,2}, v'_{k,3}, v'_{k,4})$ is the vector of updates of registers of the transition $\tau'_k$ for $k \in [6]$.

@@

At last, for the string constraints that are satisfied by a collection of strings whose lengths are small, computing LIA formulas from NFA and solving them by SMT solvers are still expensive.  
Therefore, we introduce a heuristic of searching for solutions based on \emph{under approximations}, before computing LIA formulas and solving them by SMT solvers. More precisely, we enforce a constant upper bound $B$ on the lengths of strings and search for a $B$-bounded satisfiable assignment, that is, a satisfiable assignment where the lengths of strings are bound by $B$. Initially, we set $\ell$, an integer variable to denoting\denghang{denote} the lengths of strings, to $1$. If there does not exist a \denghang{too wordy, there is no} $\ell$-bounded satisfiable assignment, then we let $\ell:=\ell+1$ and continue the search, until finding a satisfiable assignment or reaching the bound $B$. For instance, if we remove $|x| > 120$ from the string constraint~(\ref{eqn-running}), then the resulting constraint is satisfiable. Setting the bound $B$ to $3$ is sufficient to find a solution, which is witnessed by the following accepting run of $\cC$: $q^{\prime\prime}_0 \xrightarrow{(1,0,0,1)} q^{\prime\prime}_0 \xrightarrow{(0,1,0,1)} q^{\prime\prime}_1 \xrightarrow{(0,0,0,1)} q^{\prime\prime}_2$. Note that the accepting run produces $(1,1,0,3)$, a vector of register values,  which satisfies $1 \le r_1 \le 60 \wedge 1 \le r_2 \le 60 \wedge 0 \le r_3 \le 60$. (See Section~\ref{subsec:cefadec} for more details of the heuristic.)
